Thermodynamic response functions in a cell fluid model

### 2024-09-09 paper_text.tex
add footnotes to compare formulas; replace <N> with N; improve appendix

\subsection{Isothermal compressibility}
The isothermal compressibility is defined by
\begin{equation}
	\label{def:isotherm_compres}
	K_T = -\frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T,\langle N\rangle}.
\end{equation}
Let us perform some transformations to rewrite $K_T$ into a form that is more suitable for the equation of state $P=P(T,\mu)$, see~\eqref{eq:eosMT} and~\eqref{eq:eosPTM_reduced}
\begin{eqnarray*}
	K_T & = & - \frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T,\langle N\rangle}
	\\
	& = & -\rho \left(\frac{\partial \left(\frac{1}{\rho}\right)}{\partial P}\right)_{T,\langle N\rangle}
	\\
	& = & \frac{1}{\rho} \left(\frac{\partial \rho}{\partial P} \right)_{T,\langle N\rangle}
	\\
	& = & \frac{1}{\rho} \frac{\left(\partial \rho / \partial \mu\right)_{T}}
	{\left(\partial P / \partial \mu\right)_{T}},
\end{eqnarray*}
where $\rho = \langle N \rangle / V$ is the particle number density. We have omitted the condition of constant $\langle N \rangle$ in the last line of the above equation since we have explicit dependencies on temperature and chemical potential for both pressure $P = P(T, \mu)$, see Eq.~\eqref{eq:eosMT}, and density $\rho = \rho(T, \mu)$, see Eq.~\eqref{eq:density}.
Applying the Gibbs--Duhem equation
\begin{equation}
	N{\rm d}\mu = -S{\rm d}T + V{\rm d}P,
\end{equation}
at $T = const$ one has 
\begin{equation*}
	{\rm d} P = \rho {\rm d} \mu,
\end{equation*}
or
\begin{equation}
	\label{eq:rho_dpdm}
	\rho = \left(\frac{\partial P}{\partial \mu}\right)_T.
\end{equation}
Substituting this into the last expression for $K_T$, one gets
\begin{equation}
	K_T = \frac{1}{\rho^2} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
\end{equation}
Finally, from~\eqref{eq:rho_dpdm} it follows that
\begin{equation}
	\left(\frac{\partial \rho}{\partial \mu}\right)_T = \left(\frac{\partial^2 P}{\partial \mu^2}\right)_T
\end{equation}
and ultimately we arrive at the very useful expression for the isothermal compressibility
\begin{equation}
	K_T = \frac{1}{\rho^2} \left(\frac{\partial^2 P}{\partial \mu^2}\right)_{T}
\end{equation}

@@

\subsection{Thermal pressure coefficient}
The thermal pressure coefficient is defined by
\begin{equation}
	\label{def:therm_pres_coef}
	\beta_V = \left( \frac{\partial P}{\partial T} \right)_{V,\langle N \rangle}.
\end{equation}
We rewrite $\beta_V$ into a form that is suitable for the equation of state $P=P(T,\mu)$, see~\eqref{eq:eosMT} and~\eqref{eq:eosPTM_reduced}
\begin{eqnarray}
	\beta_V & = & \left(\frac{\partial P}{\partial T}\right)_{V, N}
	\nonumber\\
	& = & \left(\frac{\partial P}{\partial T}\right)_{\mu} + \left(\frac{\partial P}{\partial \mu}\right)_T \left(\frac{\partial \mu}{\partial T}\right)_{V, N}.
\end{eqnarray}
Applying the cyclic relation
\begin{equation*}
	\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
	\left(\frac{\partial T}{\partial V}\right)_{\mu, N}
	\left(\frac{\partial V}{\partial \mu}\right)_{T, N}
	= -1
\end{equation*}
we obtain
\begin{eqnarray*}
	\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
	& = & 
	- \left(\frac{\partial V}{\partial T}\right)_{\mu, N} 
	\left(\frac{\partial V}{\partial \mu}\right)^{-1}_{T, N} \\
	& = & - \left(\frac{\partial \bar{n}}{\partial T}\right)_{\mu}
	\left(\frac{\partial \bar{n}}{\partial \mu}\right)^{-1}_{T}.		
\end{eqnarray*}
Substituting this result into the formula for $\beta_V$, we arrive at the final expression for the thermal pressure coefficient
\begin{equation}
	\beta_V = \left(\frac{\partial P}{\partial T}\right)_{\mu} 
	- \left(\frac{\partial P}{\partial \mu}\right)_T 
	\left(\frac{\partial \bar{n}}{\partial T}\right)_{\mu}
	\left(\frac{\partial \bar{n}}{\partial \mu}\right)^{-1}_{T}
\end{equation}
which is easily calculated based on Eqs.~\eqref{eq:eosMT} and~\eqref{eq:density}.\\
(NB. Compare this equation with Eq.(17) from~\cite{StrokerMeier2021})

@@

\subsection{Thermal expansion coefficient}
The thermal expansion coefficient is defined by
\begin{equation}
	\alpha_P = \frac{1}{V}\left(\frac{\partial V}{\partial T}\right)_{P,\langle N\rangle}.
\end{equation}
We rewrite $\alpha_P$ into a form suitable for the equation of state $P=P(T,\mu)$, see~\eqref{eq:eosMT} and~\eqref{eq:eosPTM_reduced}
\begin{eqnarray}
	\alpha_P & = & \frac{1}{V}\left(\frac{\partial V}{\partial T}\right)_{P, N}
	\nonumber\\
	\label{eq:alpha_vs_rho}
	& = & -\frac{1}{\rho} \left(\frac{\partial \rho}{\partial T}\right)_{P, N}
	\\
	& = & -\frac{1}{\rho} \left(\frac{\partial \rho}{\partial T}\right)_{\mu} 
	- \frac{1}{\rho} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
	\left(\frac{\partial \mu}{\partial T}\right)_{P, N}.
\end{eqnarray}
Applying the cyclic relation
\begin{equation*}
	\left(\frac{\partial \mu}{\partial T}\right)_{P, N}
	\left(\frac{\partial T}{\partial P}\right)_{\mu, N}
	\left(\frac{\partial P}{\partial \mu}\right)_{T, N}
	= -1
\end{equation*}
we get
\begin{eqnarray*}
	\left(\frac{\partial \mu}{\partial T}\right)_{P, N} & = & 
	- \left(\frac{\partial P}{\partial T}\right)_{\mu}
	\left(\frac{\partial P}{\partial \mu}\right)^{-1}_{T}.
\end{eqnarray*}
Substituting the result into the expression for $\alpha_P$, we arrive at the final formula for the thermal expansion coefficient
\begin{equation}
	\alpha_P = -\frac{1}{\rho}\left(\frac{\partial \rho}{\partial T}\right)_{\mu}
	+ \frac{1}{\rho} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
	\left(\frac{\partial P}{\partial T}\right)_{\mu}
	\left(\frac{\partial P}{\partial \mu}\right)^{-1}_{T}
\end{equation}
which is readily calculated based on Eqs.~\eqref{eq:eosMT} and~\eqref{eq:density}.\\
(NB. Compare this equation with Eqs.(57)-(58) from~\cite{StrokerMeier2021})

We introduce the reduced thermal expansion coefficient by
\begin{equation}
	\alpha^*_P = \frac{\varepsilon}{k_{\rm B}} \alpha_P.
\end{equation}
It is expressed in terms of the reduced quantities $P^*$ and $\bar{n}$ as follows
\begin{equation}
	\alpha^*_P = \frac{1}{T^*_c}
	\left( 
	-\frac{1}{\bar{n}}\left(\frac{\partial \bar{n}}{\partial \tau}\right)_{\mu}
	+ \frac{1}{\bar{n}} \left(\frac{\partial \bar{n}}{\partial \mu}\right)_{T}
	\left(\frac{\partial P^*}{\partial \tau}\right)_{\mu}
	\left(\frac{\partial P^*}{\partial \mu}\right)^{-1}_{T} 
	\right).
\end{equation}

@@

\section{\label{sec:app-a} Explicit expressions for derivatives}
In this appendix, explicit expressions are presented for derivatives of quantities entering the equation of state, with respect to temperature. This expressions are useful while calculating the response functions in Section~\ref{sec:respons_functions}.

>>>>

\subsection{Isothermal compressibility}
The isothermal compressibility is defined by
\begin{equation}
	\label{def:isotherm_compres}
	K_T = -\frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T, N}.
\end{equation}
Let us perform some transformations to rewrite $K_T$ into a form that is more suitable for the equation of state $P=P(T,\mu)$, see~\eqref{eq:eosMT} and~\eqref{eq:eosPTM_reduced}
\begin{eqnarray*}
	K_T & = & - \frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T,N}
	\\
	& = & \frac{1}{\rho} \left(\frac{\partial \rho}{\partial P} \right)_{T,N}
	\\
	& = & \frac{1}{\rho} \frac{\left(\partial \rho / \partial \mu\right)_{T}}
	{\left(\partial P / \partial \mu\right)_{T}},
\end{eqnarray*}
where $\rho = \langle N \rangle / V$ is the particle number density. We have omitted the condition of constant $\langle N \rangle$ in the last line of the above equation since we have explicit dependencies on temperature and chemical potential for both pressure $P = P(T, \mu)$, see Eq.~\eqref{eq:eosMT}, and density $\rho = \rho(T, \mu)$, see Eq.~\eqref{eq:density}.
Applying the Gibbs--Duhem equation
\begin{equation}
	N{\rm d}\mu = -S{\rm d}T + V{\rm d}P,
\end{equation}
at $T = const$ one has 
\begin{equation*}
	{\rm d} P = \rho {\rm d} \mu,
\end{equation*}
or
\begin{equation}
	\label{eq:rho_dpdm}
	\rho = \left(\frac{\partial P}{\partial \mu}\right)_T.
\end{equation}
Substituting this into the last expression for $K_T$, one gets
\begin{equation}
	K_T = \frac{1}{\rho^2} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
\end{equation}
Finally, from~\eqref{eq:rho_dpdm} it follows that
\begin{equation}
	\left(\frac{\partial \rho}{\partial \mu}\right)_T = \left(\frac{\partial^2 P}{\partial \mu^2}\right)_T
\end{equation}
and ultimately we arrive at the very useful expression for the isothermal compressibility
\begin{equation}
	K_T = \frac{1}{\rho^2} \left(\frac{\partial^2 P}{\partial \mu^2}\right)_{T}
\end{equation}

@@

\subsection{Thermal pressure coefficient}
The thermal pressure coefficient is defined by
\begin{equation}
	\label{def:therm_pres_coef}
	\beta_V = \left( \frac{\partial P}{\partial T} \right)_{V,\langle N \rangle}.
\end{equation}
We rewrite $\beta_V$ into a form that is suitable for the equation of state $P=P(T,\mu)$, see~\eqref{eq:eosMT} and~\eqref{eq:eosPTM_reduced}
\begin{eqnarray}
	\beta_V & = & \left(\frac{\partial P}{\partial T}\right)_{V, N}
	\nonumber\\
	& = & \left(\frac{\partial P}{\partial T}\right)_{\mu} + \left(\frac{\partial P}{\partial \mu}\right)_T \left(\frac{\partial \mu}{\partial T}\right)_{V, N}.
\end{eqnarray}
Applying the cyclic relation
\begin{equation*}
	\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
	\left(\frac{\partial T}{\partial V}\right)_{\mu, N}
	\left(\frac{\partial V}{\partial \mu}\right)_{T, N}
	= -1
\end{equation*}
we obtain
\begin{eqnarray*}
	\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
	& = & 
	- \left(\frac{\partial V}{\partial T}\right)_{\mu, N} 
	\left(\frac{\partial V}{\partial \mu}\right)^{-1}_{T, N} \\
	& = & - \left(\frac{\partial \bar{n}}{\partial T}\right)_{\mu}
	\left(\frac{\partial \bar{n}}{\partial \mu}\right)^{-1}_{T}.		
\end{eqnarray*}
Substituting this result into the formula for $\beta_V$, we arrive at the final expression for the thermal pressure coefficient
\begin{equation}
	\beta_V = \left(\frac{\partial P}{\partial T}\right)_{\mu} 
	- \left(\frac{\partial P}{\partial \mu}\right)_T 
	\left(\frac{\partial \bar{n}}{\partial T}\right)_{\mu}
	\left(\frac{\partial \bar{n}}{\partial \mu}\right)^{-1}_{T}
\end{equation}
which\footnote{Compare this equation for $\beta_V$ with Eq.(17) from~\cite{StrokerMeier2021}} is easily calculated based on Eqs.~\eqref{eq:eosMT} and~\eqref{eq:density}.

@@

\subsection{Thermal expansion coefficient}
The thermal expansion coefficient is defined by
\begin{equation}
	\alpha_P = \frac{1}{V}\left(\frac{\partial V}{\partial T}\right)_{P,N}.
\end{equation}
We rewrite $\alpha_P$ into a form suitable for the equation of state $P=P(T,\mu)$, see~\eqref{eq:eosMT} and~\eqref{eq:eosPTM_reduced}
\begin{eqnarray}
	\alpha_P & = & \frac{1}{V}\left(\frac{\partial V}{\partial T}\right)_{P, N}
	\nonumber\\
	\label{eq:alpha_vs_rho}
	& = & -\frac{1}{\rho} \left(\frac{\partial \rho}{\partial T}\right)_{P, N}
	\\
	& = & -\frac{1}{\rho} \left(\frac{\partial \rho}{\partial T}\right)_{\mu} 
	- \frac{1}{\rho} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
	\left(\frac{\partial \mu}{\partial T}\right)_{P, N}.
\end{eqnarray}
Applying the cyclic relation
\begin{equation*}
	\left(\frac{\partial \mu}{\partial T}\right)_{P, N}
	\left(\frac{\partial T}{\partial P}\right)_{\mu, N}
	\left(\frac{\partial P}{\partial \mu}\right)_{T, N}
	= -1
\end{equation*}
we get
\begin{eqnarray*}
	\left(\frac{\partial \mu}{\partial T}\right)_{P, N} & = & 
	- \left(\frac{\partial P}{\partial T}\right)_{\mu}
	\left(\frac{\partial P}{\partial \mu}\right)^{-1}_{T}.
\end{eqnarray*}
Substituting the result into the expression for $\alpha_P$, we arrive at the final formula for the thermal expansion coefficient
\begin{equation}
	\label{eq:alphaP1}
	\alpha_P = -\frac{1}{\rho}\left(\frac{\partial \rho}{\partial T}\right)_{\mu}
	+ \frac{1}{\rho} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
	\left(\frac{\partial P}{\partial T}\right)_{\mu}
	\left(\frac{\partial P}{\partial \mu}\right)^{-1}_{T}
\end{equation}
which\footnote{Compare this equation for $\alpha_P$ with Eqs.(57)-(58) from~\cite{StrokerMeier2021}} is readily calculated based on Eqs.~\eqref{eq:eosMT} and~\eqref{eq:density}.

We introduce the reduced thermal expansion coefficient by
\begin{equation}
	\alpha^*_P = \frac{\varepsilon}{k_{\rm B}} \alpha_P.
\end{equation}
It is expressed in terms of the reduced quantities $P^*$ and $\bar{n}$ as follows
\begin{equation}
	\alpha^*_P = \frac{1}{T^*_c \bar{n}}
	\left[ 
	-\left(\frac{\partial \bar{n}}{\partial \tau}\right)_{\mu}
	+ \left(\frac{\partial \bar{n}}{\partial \mu}\right)_{T}
	\left(\frac{\partial P^*}{\partial \tau}\right)_{\mu}
	\left(\frac{\partial P^*}{\partial \mu}\right)^{-1}_{T} 
	\right].
\end{equation}

@@

\section{\label{sec:app-a} Explicit expressions for derivatives}
In this Appendix, we provide an explicit example of differentiating the equation of state, namely~\eqref{eq:eosPTn_reduced}, by deriving the expression for the thermal pressure coefficient $\beta^*_V$ from~\eqref{eq:beta_star_n}.
